---
title: "Basic Probability"
date: "March 17th, 2016"
layout: post
output: 
  html_document:
tags: [R, stats]
---

Probabilites represent the chances of an event occurring. Given a random process (e.g. flipping a coin), probabilities can be numerically quantified by measuring the number of desired outcomes divided by the total number of outcomes. For instance, if we flip a coin the two possible outcomes are head or tails. The probability of getting a head is then 1/2. 

Probabilites may be either marginal, joint or conditional.

```{r echo = FALSE}
library("knitr")
```

## Joint Probabilities



Joint probabilities are

```{r, message = FALSE}
library("ggplot2")
library("dplyr")
library("reshape2")
library("knitr")

diamonds.prop.df <- 
  diamonds %>%
  group_by(color, cut) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(prop = n / sum(n))

diamonds.prop.df %>%
  dcast(color ~ cut, value.nar = "prop") %>%
  kable(align = "l")
```

## Marginal Probabilities

This is simply the probability of some event a occuring. For instance, the probability of a coin flip giving heads is p(a) = 0.5.

```{r}
color.marginal.df <- 
  diamonds.prop.df %>%
  group_by(color) %>%
  summarize(marginal = sum(prop))

kable(color.marginal.df, align = "l")
```

```{r}
cut.marginal.df <- 
  diamonds.prop.df %>%
  group_by(cut) %>%
  summarize(marginal = sum(prop))

kable(cut.marginal.df, align = "l")
```

## Conditional Probability



## References

* [Nicholas School Statistics Review - Joint, Marginal and Conditional Probabilities](http://sites.nicholas.duke.edu/statsreview/probability/jmc/)

