---
title: "Bayes' Rule"
date: "March 17th, 2016"
layout: post
output:
  html_document:
tags: [R, stats]
---

  
Here is an overview of what will be discussed in this post.

**Table of Contents**

<ul data-toc="body" data-toc-headings="h2,h3"></ul>

```{r echo = FALSE}
library("knitr")
library("captioner")

tbl.nums <- captioner(prefix = "<u>Table</u>", css_class = "tblcaption")
```

In my previous post,

Bayes rules can be derived by starting with the conditional probability of <span class="inlinecode">$P(X\ |\ Y)$</span> which is represented as:

<div>
$$P(X\ |\ Y) = \frac{P(X, Y)}{P(Y)}$$
</div>

First, let's start by multiplying this with P(Y():

<div>
$$P(X\ |\ Y)\ P(Y) = P(X, Y)$$
</div>

We can also perform the algebraic manipulation for <span class="inlinecode">$P(Y\ |\ X)$</span>:

<div>
$$\begin{align}
P(Y\ |\ X) &= \frac{P(X, Y)}{P(X)} \\
P(Y\ |\ X)\ P(X) &= P(X, Y)
\end{align}$$
</div>

Notice how we now have two alternative representations of the joint probability <span class="inlinecode">$P(X, Y)$</span> which we can equate to each other:

<div>
$$\begin{align}
P(X\ |\ Y)\ P(Y) &= P(X, Y) \\
P(Y\ |\ X)\ P(X) &= P(X, Y) \\
P(X\ |\ Y)\ P(Y) &= P(Y\ |\ X)\ P(X) 
\end{align}$$
</div>

If we now have divide by <span class="inlinecode">$P(Y)$</span>:

<div>
$$P(X\ |\ Y) = \frac{P(Y\ |\ X)\ P(X)}{P(Y)}$$
</div>

This is defined as the Bayes' rule.

## Bayes' Rule and the Two-Way Table

In my [Joint, Marginal, and Conditional Probabilities post]({% post_url 2016-03-20-basic-prob %}), we produced a two-way joiny probability table on diamond color-cut:

```{r, message = FALSE}
library("ggplot2")
library("dplyr")
library("reshape2")
library("knitr")

diamonds.color.cut.df <-
  diamonds %>%
  group_by(color, cut) %>%
  summarize(n = n())

diamonds.color.cut.prop.df <- 
  diamonds.color.cut.df %>%
  ungroup() %>%
  mutate(prop = n / sum(n))

diamonds.color.cut.prop.df %>%
  dcast(color ~ cut, value.var = "prop") %>%
  kable(align = "l", format = "html", 
        table.attr = 'class="table table-striped table-hover"')
```

## References

## R Session

```{r session, echo = FALSE}
devtools::session_info()
```
