---
title: "Fitting a Binomial Mixture Model Using the Expectation-Maximization Algorithm in R"
date: "January 3rd, 2016"
layout: post
output: 
  html_document:
    toc: true
tags: [R, mixmodels, EM, bmm]
---

```{r echo = FALSE}
library("knitr")
library("svglite")
knitr::opts_chunk$set(fig.path="{{ site.url }}/assets/bmm-em/")
                      #dev = "svglite")
```

In my previous post ["Fitting a Gaussian Mixture Model Using the Expectation-Maximization Algorithm in R"]({% post_url 2016-01-03-gmm-em %}), I went over how to implement a mixture model using the expectation-maximization (EM) algorithm in R. Specifically, this was done with gaussian distributions as our components. In this post, I will show how we can use binomial distributions instead of gaussian distributions allowing us to perform binomial mixture model (BMM) clustering.

**Table of Contents**

<ul data-toc="body" data-toc-headings="h2,h3"></ul>

## Setup

Let's setup a scenario of where a BMM would be useful. Let's say for example, you have a bag that has 2 coins in it. You run experiment where you do the following:

1. Reach your hand into the bag and draw one of these coins. 
1. You then proceed to flip this coin 60 times and note the number of heads.
1. Put the coin back in the bag.
1. Repeats steps 1-3 a total of 1000 times.

Say you ran this very very time consuming experiment and these were your results:

```{r bmm-scenario, message = FALSE}
library("dplyr")
library("ggplot2")

set.seed(1)

#' Random BMM data generator
#'
#' Code is taken from https://cran.r-project.org/web/packages/sBIC/vignettes/BinomialMixtures.pdf
#' @param n Number of experiments.
#' @param alpha Mixing weights of the components.
#' @param theta Probability of success of each component.
#' @param size Number of trials in each experiment.
#' @return vector Data from a BMM
rbinommix <- function(n, alpha, theta, size) {
  nindex = rmultinom(1, size = n, prob = alpha)
  rbinom(n, size, rep(theta, nindex))
}

num.trials <- 60

coin.flip.data <- 
  rbinommix(1000, 
            alpha = c(0.4, 0.6), 
            theta = c(0.35, 0.65), 
            size = num.trials)

data_frame(num_successes = coin.flip.data) %>%
  ggplot(aes(x = num_successes)) +
  geom_histogram(binwidth = 1) +
  xlab("Number of Heads") +
  ylab("Number of Experiments")
```

Fill in details...


## The Binomial Distribution

Recall from the ["Using Mixture Models for Clustering in R"]({% post_url 2015-10-13-mixture-model %}) post, a mixture model is a mixture of <span class="inlinecode">$k$</span> component distributions that collectively make a mixture distribution <span class="inlinecode">$f(x)$</span>:

<div>
$$f(x) = \sum_{k=1}^{K}\alpha_{k}f_{k}(x)$$
</div>

The <span class="inlinecode">$\alpha_{k}$</span> represents a mixing weight for the <span class="inlinecode">$k^{th}$</span> component where <span class="inlinecode">$\sum_{k=1}^{K}\alpha_{k} = 1$</span>. Remember that the <span class="inlinecode">$f_k(x)$</span> components in principle are arbitrary in the sense that you can choose any sort of distribution. In this post, our components will be binomial distributions to form a BMM. A binomial distribution is a discrete probability distribution that is represented as:

<div>
$$
B(n, p)
$$
</div>

The binomial distribution is parameterized by two variables:

* <span class="inlinecode">$n$</span>: Number of trials.
* <span class="inlinecode">$p \in [0, 1]$</span>: Success probability in each trial.

The distribution is often used to model the number of "successes" in a sample size of n drawn with replacement from a population of size N. The modeling of coin flips is often an example of where the binomial distribution is used. This is because each coin flip has a probability of giving a head (assuming head is what we define as a "success"). We can flip the coin n number of times to generate a sample size of n. Each coin flip is independent of other coin flips and thus is ....TO FILL IN.

## Fitting a BMM using Expectation Maximization 

Recall that the EM algorithm consists of 3 major steps:

1. Initialization
1. Expectation (E-step)
1. Maximization (M-step)

Steps 2 and 3 are repeated until convergence. We will cover each of these steps and how convergence is reached below. But first we must understand how to mathematically represent a BMM:

<div>
$$P(X\ |\ n,p,\alpha) = \sum_{k=1}^{K}\alpha_kB(X\ |\ n_{k},p_{k})$$
</div>

* X = Dataset of n elements (<span class="inlinecode">$x_{1}, ..., x_{n}$</span>).
* <span class="inlinecode">$\alpha_{k}$</span> = Mixing weight of the <span class="inlinecode">k</span>th component. <span class="inlinecode">$\sum_{k=1}^{K}\alpha_{k} = 1$</span>.
* <span class="inlinecode">$B(X\ |\ n_{k},p_{k})$</span> = [Binomial probability mass function (pmf)](https://en.wikipedia.org/wiki/Binomial_distribution) of the <span class="inlinecode">k</span>th component defined by the parameters <span class="inlinecode">$n$</span> and <span class="inlinecode">$p$</span>.
* <span class="inlinecode">$n_{k}$</span> = Number of trials for each experiment in the <span class="inlinecode">$k$</span>th component. In this scenario, this number is 60 for every experiment.
* <span class="inlinecode">$p_{k}$</span> = Probability of a success in the <span class="inlinecode">$k$</span>th component.

So for a two component BMM, we would mathematically represent this as:

<div>
$$P(X\ |\ n,p,\alpha) = \alpha_{1}B(X\ |\ n_{1}, p_{1}) + \alpha_{2}B(X\ |\ n_{2},p_{2})$$
</div>

In case you were wondering what the <span class="inlinecode">$P(X|\mu,\sigma,\alpha)$</span> means, don't worry about that for now. We will explain exactly what this means later on in the post.

### Initialization: Determining the Initial BMM Parameters

So the first thing we need to do is to set the **initial model parameters** of a BMM. I am not entirely sure if there is a standard way of doing this, but k-means should be a viable approach to performing some initial cluster and getting the initial BMM parameters from these clusters.

* <span class="inlinecode">$p_{k} = \frac{\sum_{i}^{N_{k}}(x_{i,k} - \mu_{k})^2}{N_{k}}$</span>
* <span class="inlinecode">$\alpha_{k} = \frac{N_{k}}{N}$</span>

Where <span class="inlinecode">$N_{k}$</span> indicates the number of data points in the kth component. Let's try that here:

```{r kmeans_init, fig.height = 4}
#coin.flip.data.kmeans <- kmeans(coin.flip.data, 2)
#
#coin.flip.data.kmeans.cluster <- coin.flip.data.kmeans$cluster

coin.flip.data.kmeans.df <- 
  data_frame(num_heads = coin.flip.data, 
             cluster = base::sample(1:2, 
                                    length(coin.flip.data), 
                                    replace = TRUE))

coin.flip.data.kmeans.df %>%
  ggplot(aes(x = num_heads, fill = factor(cluster))) +
  geom_density(alpha = 0.4) +
  xlab("Number of Heads") +
  ylab("Density") +
  scale_color_discrete(name = "Cluster") +
  ggtitle("K-means Clustering") +
  scale_fill_discrete(name = "Cluster")
```

Since we specified 2 clusters, k-means nicely splits the data into 2 clusters. We can calculate the <span class="inlinecode">$p$</span> and <span class="inlinecode">$\alpha$</span>

```{r}
coin.flip.data.kmeans.summary.df <- 
  coin.flip.data.kmeans.df %>%
  mutate(prob = num_heads / num.trials) %>%
  group_by(cluster) %>%
  summarize(prob = mean(prob), size = n()) %>%
  mutate(alpha = size / sum(size))

kable(coin.flip.data.kmeans.summary.df,
      caption = "Initial parameters of the BMM")
```

### Expectation: Calculating the "Soft Labels" of Each Data Point (E-step)

Now that we have the initial parameters of our BMM, we now have to determine what is the probability (soft label; responsibility) that the data point (<span class="inlinecode">$x_{i}$</span>) belongs to component (<span class="inlinecode">$k_{j}$</span>)? This is considered the expectation step (E-step) of MLE where we are calculating the "expectation values" of the soft labels for each data point. 

Mathematically, the question can be posed like this <span class="inlinecode">$P(x_{i} \in k_{j} | x_{i})$</span>. How do we actually solve this equation? To help us, we can apply [Bayes' rule](https://en.wikipedia.org/wiki/Bayes%27_rule) here:

<div>
$$P(x_{i} \in k_{j} | x_{i}) = \frac{P(x_{i} | x_{i} \in k_{j})\ P(k_{j})}{P(x_{i})}$$
</div>

The parts of this equation are related to the BMM equation above as follows:

* <span class="inlinecode">$P(x_{i} | x_{i} \in k_{j}) = B(x_{i}\ |\ n,p_{k_{j}})$</span>
* <span class="inlinecode">$P(k_{j}) = \alpha_{k_{j}}$</span>
* <span class="inlinecode">$P(x_{i}) = \sum_{k=1}^{K}\alpha_{k}B(x_{i}\ |\ n,p_{k_{j}})$</span>

What we are interested in is <span class="inlinecode">$P(x_{i} \in k_{j} | x_{i})$</span> which is called the posterior probability. Knowing these equations, we can easily calculate this. For instance, what is the posterior probability of x = 66 belong to the first component? We can first calculate the top part of the equation like this in R:

```{r}
num.heads <- 30
coin.flip.data.kmeans.summary.df

comp1.prod <- 
  dbinom(num.heads, num.trials, coin.flip.data.kmeans.summary.df$prob[1]) *
  coin.flip.data.kmeans.summary.df$alpha[1]
```

Here we are using the `binom` function from R to make use of the binomial pmf To calculate the bottom part of the equation, we actually need to calculate this value for both components and sum them up:

```{r}
comp2.prod <- 
  dbinom(num.heads, num.trials, coin.flip.data.kmeans.summary.df$prob[2]) *
  coin.flip.data.kmeans.summary.df$alpha[2]

normalizer <- comp1.prod + comp2.prod
```

Now that we have all the components of the equation, let's plug and solve this:

```{r}
comp1.prod / normalizer
```

We can easily calculate this for every data point as follows:

```{r}
comp1.prod <- dbinom(x = coin.flip.data, 
                     size = num.trials,
                     prob = coin.flip.data.kmeans.summary.df$prob[1])

comp2.prod <- dbinom(x = coin.flip.data, 
                     size = num.trials,
                     prob = coin.flip.data.kmeans.summary.df$prob[2])

normalizer <- comp1.prod + comp2.prod

comp1.post <- comp1.prod / normalizer
comp2.post <- comp2.prod / normalizer
```

### Maximization: Re-estimate the Component Parameters (M-step)

Now that we have posterior probabilites (i.e. soft labels), we can re-estimate our component parameters. We simply have to make a little adjustment to the MLE equations that we specified early. Specifically, the <span class="inlinecode">$N_{k}$</span> (remember there are no hard labels) is replaced with the posterior probability <span class="inlinecode">$P(x_{i} \in k_{j} | x_{i})$</span> in each equation.

* <span class="inlinecode">$\mu_{k} = \frac{\sum_{i}^{N}P(x_{i} \in k_{j} | x_{i})x_{i}}{\sum_{i}^{N}P(x_{i} \in k_{j} | x_{i})}$</span>
* <span class="inlinecode">$\sigma_{k} = \frac{\sum_{i}^{N}P(x_{i} \in k_{j} | x_{i})(x_{i} - \mu_{k})^2}{\sum_{i}^{N}P(x_{i} \in k_{j} | x_{i})}$</span>
* <span class="inlinecode">$\alpha_{k} = \frac{\sum_{i}^{N}P(x_{i} \in k_{j} | x_{i})}{N}$</span>

With these equations we can now plug in our values and calculate the components parameters using our example from above:

```{r}
comp1.n <- sum(comp1.post)
comp2.n <- sum(comp2.post)

comp1.prob <- 1/comp1.n * sum(comp1.post * coin.flip.data)
comp2.prob <- 1/comp2.n * sum(comp2.post * coin.flip.data)

comp1.alpha <- comp1.n / length(coin.flip.data)
comp2.alpha <- comp2.n / length(coin.flip.data)

comp.params.df <- data.frame(comp = c("comp1", "comp2"),
                             comp.prob = c(comp1.prob, comp2.prob),
                             comp.alpha = c(comp1.alpha, comp2.alpha),
                             comp.cal = c("self", "self"))
comp.params.df
```


## Summary

## R Session

```{r}
devtools::session_info()
```

## References

* [Youtube Videos on Mixture Models](https://www.youtube.com/playlist?list=PLBv09BD7ez_4e9LtmK626Evn1ion6ynrt)
* [Mixture Models](http://www.stat.cmu.edu/~cshalizi/uADA/12/lectures/ch20.pdf)
* [Expectation Maximization and Gaussian Mixture Models](http://www.slideshare.net/petitegeek/expectation-maximization-and-gaussian-mixture-models)
* [Nature Computational Biology Primer - What is the expectation maximization algorithm?](http://www.nature.com/nbt/journal/v26/n8/full/nbt1406.html)
Thankfully we have co
