---
title: "The Basics of Survival Analysis"
date: "April 21st, 2016"
layout: post
output:
  html_document
tags: [R, stats, bayesian]
---



Here is an overview of what will be discussed in this post.

**Table of Contents**

<ul data-toc="body" data-toc-headings="h2,h3"></ul>

```{r echo = FALSE}
library("knitr")
library("captioner")

tbl.nums <- captioner(prefix = "<u>Table</u>", css_class = "tblcaption")
```

Survivor functions (curves) can be estimated by using the Kaplan-Meier method. Here in this post, we will use the colon dataset from the survival R package.

```{r}
library("survival")
library("dplyr")
library("reshape2")

colon %>%
  filter(rx == "Obs")
```

The help understand how the Kaplan-Meier method works, it's best to represent the survival data in a particular tabular format:

| Ordered Event Times | No. Events at Mf | No. Censored 

```{r}
fail.times <- 
  colon %>%
  filter(rx == "Obs", status == 1) %>% 
  distinct(time) %>%
  arrange(time) %>%
  .$time

colon %>%
  filter(rx == "Obs", time > 0, time <= 20)

fail.times

  group_by(time, status) %>%
  summarize(num_patients = n()) %>%
  dcast
  
  

```



# References

* [Survival Analysis - A Self Learning Text](http://www.amazon.ca/Survival-Analysis-Statistics-Biology-Health-ebook/dp/B00DGEF822?ie=UTF8&qid=&ref_=tmm_kin_swatch_0&sr=)
