

# Create Test Data

Here we create some linear relationship between x and y. But we add some noise in the y-value.

```{r}
trueA <- 5
trueB <- 0 # y-intercept
trueSd <- 10
sampleSize <- 31
 
# create independent x-values 
x <- (-(sampleSize-1)/2):((sampleSize-1)/2)

# create dependent values according to ax + b + N(0,sd)
y <-  trueA * x + trueB + rnorm(n = sampleSize, mean = 0, sd = trueSd)
 
plot(x,y, main="Test Data")
```

# Defining the statistical model
***

In machine learning, there are 3 elements to any model:

1. Representation (Hypothesis Function)
2. Evaluation (Loss Function; Objective Function)
3. Optimization

The representation of our statistical model will be:

$$Y = aX + b + N(0, sd)$$

# Defining the Likelihood (Loss Function)
***

We define our likelihood function (aka. loss/objective function). In brief, a likelihood function is:

> probability (density) with which we would expect the observed data to occur conditional on the parameters of the model that we look at. 

The loss function of a traditional simple linear regression is:

$$L = \sum_{i=1}^{n}(Y_{i} - (\beta_{0} + \beta_{1}X_{i}))^2$$

```{r}
#' @param param Vector of length x containing slope, y-intercept, and standard
#'   deviation.
likelihood <- function(param) {

	a <- param[1]
	b <- param[2]

	# param[3] = trueSd
	in.sd <- param[3]
		
	# Make predictions given slope (a)
	# NB: x is fixed data
	# NB: b == trueB
	pred <- a*x + b

	# log is used to get log likelihoods
	# For each pred, we run dnorm with the pred as a mean, in.sd (fixed across
	# dnorm runs) with the corresponding y value. 
	#
	# The idea is to get the probability density of observing the deviation of
	# prediction to the true value. We then sum up the probability density values
	# across all of these predictions.
	singlelikelihoods <- dnorm(y, mean = pred, sd = in.sd, log = TRUE)
	sumll <- sum(singlelikelihoods)
	sumll
}

# Example: plot the likelihood profile of the slope a

#' @param x Slope of the line
slopevalues <- function(in.slope) {
	# trueB is fixed
	# trueSd is fixed
	likelihood(param = c(in.slope, trueB, trueSd))
}

# Produce a vector of slopes
slope <- seq(3, 7, by = .05)
in.slope <- 3

# Run slopevalues for each slope value
slopelikelihoods <- lapply(slope, slopevalues)

plot(seq(3, 7, by=.05), slopelikelihoods , type = "l", 
		 xlab = "values of slope parameter a", 
		 ylab = "Log likelihood")
```

Our stati
