---
title: "Power and Sample Size Calculations"
date: "March 17th, 2016"
layout: post
output:
  html_document
tags: [R, stats]
---

* In the context of hypothesis testing.
    + Null hypothesis and alternative hypothesis.
  * Need to define the crticial value. <span class="inlinecode">$alpha$</span>
* Alpha level: Probability of rejecting the null hypothesis assuming that the null hypothesis is true.
* Statistical power: Probability of rejecting the null hypothesis assuming that the alternative hypothesis is true. 
    + 1 - P(Type II error)
* Well-designed test will have high statistical power (e.g. 0.8 and 0.9)
    + We can increase power by increasing the effect size which is the difference between the parameter values of our null and alternative hypothesis
    + Increase sample size. This will reduce our standard error.
* To solve for power, we need 4 things:
    1. Alpha
        + one-sided or two sides
        + typically 0.05 or 0.01
    1. Sample size: Generally dependent on budget and logical constraints
    1. Effect size: Difference between the hypothesized parameter values
    1. Variance: Based on previous studies or pilot data.


    Sometimes the power is not something we get to selected as it is dicated by the funding agency
* We can solve the minimum detectable effect size.
* If we calculate power, then we have a specific sample-size.
* If we calculate sample-size, then we have a specific power we are interested.

Here is an overview of what will be discussed in this post.

**Table of Contents**

<ul data-toc="body" data-toc-headings="h2,h3"></ul>

```{r echo = FALSE}
library("knitr")
library("captioner")

tbl.nums <- captioner(prefix = "<u>Table</u>", css_class = "tblcaption")
```

## Power Analysis in R

We can make use of the [pwr R package](https://cran.r-project.org/web/packages/pwr/index.html) to do some power calculations. For instance, we can use the `pwr.t2n.test` to do a power calculation for a two-sample t-test where the two groups have unequal sample sizes. For every test in this package, as long as we have 3 of the 4 following quantities we can calculate the missing quantity:

1. Sample size
1. Effect size
1. Significance level
1. Power

For instance:

```{r }
library("pwr")
pwr.t2n.test(n1 = 30, n2 = 270, sig.level = 0.05, power = 0.9)
```

For two-sample t-tests, the effect sizes are measured as

<div>
$$d = \frac{|\mu_{1}-\mu_{2}}{\sigma}$$
</div>

Cohen suggests that d values of 0.2, 0.5, and 0.8 represent small, medium, and large effect sizes respectively.

## References

* [Inference for Means: Comparing Two Independent Samples](http://www.stat.ubc.ca/~rollin/stats/ssize/n2.html)
* [A conceptual introduction to power and sample size calculations using Stata](https://www.youtube.com/watch?v=QBONLUp7i28)
* [Quick-R: Power Analysis](http://www.statmethods.net/stats/power.html)
* [idre - UCLA: Power Analysis for Two-group Independent sample t-test](http://www.ats.ucla.edu/stat/r/dae/t_test_power2.htm)

## R Session

```{r session, echo = FALSE}
devtools::session_info()
```
