---
title: "What are Probability Distributions"
layout: post
output: 
  html_document:
    toc: true
tags: []
---

```{r echo = FALSE}
library("knitr")
#knitr::opts_chunk$set(fig.path="{{ site.url }}/assets/gmm-em/")
```

## What is a Random Variable?

A random variable can be a confusing concept because it is not like a traditional variable that you may have been exposed to before. A random variable are actually functions that map the outcomes of a random process to a numeric value. For instance, let's say we have the random process of flipping a coin and we use the random variable X (random variables are typically denoted with capital letters) to represent the outcomes of the coin flip. For example, here is one possible random variable:

* X = 1 if the flip of the coin is a head
* X = 0 if the flip of the coin is a tail

There isn't anything special about the numeric values I chose here. I could have done the following:

* X = 500 if the flip of the coin is a head
* X = 25 if the flip of the coin is a tail

And this would still be e a random variable. Note how the flipping of a coin is a random process. This is because each flip of the coin can produce either a heads or a tail and is essentially a random event. If we had some weird coin that had both sides being head, then the flipping of this coin would be a deterministic process since we would always get a head every flip.

Random variables allows for easier mathematical notation. 

### Traditional vs. Random Variable

* traditional variables are typically represented by lowercase letters (e.g. x, y)
* traditional variables can be solved for (e.g. x + 10 = 15); we don't solve for random variables
* traditional variables can be used to observed how a value changes as function of the variable (e.g. y = x + 2)
* random variables can take on many different values, and we are often interested in knowing what the probability of a random variable taking on a specific value.

* Easier notation

Probability distributions might be something you've heard of it, sort of understood, but **never** really understood what it was. I will admit that in my work I often used the term, wrote about it, but when it came down to describing it I can honestly say I didn't fully understand it. In this blog post, I hope to be able to fully explain what a probability distribution is.

## So What is a Probability Distribution?

A probability distribution (aka. probability distribution table) is a way to represent all possible outcomes and their respective mutually exclusive probabilities. For instance, imagine you had a coin that had two sides (heads and tails) with equal chance of getting a head or tail on a given flip. Well then the probability distribution is then:

| Outcomes | Probability |
|:--------:|:-----------:|
| H        | 1/2         |
| T        | 1/2         |

We have all possible mutually exclusive outcomes in the "Outcomes" column and their associated probabilities in the "Probability" column. Additionally, the probabilites satisfy 3 fundamental properties of probabilty:

1. Value has to be non-negative (can't have negative probability).
1. Sum of probabilites across all possible outcomes must equal 1.
1. For two mutually exclusive events, the probability of one or the other is the sum of their individual properties.

Let's use a more interesting experiment where instead of flipping a coin once, we flip the coin 3 times. In our case, we let X represent the number of heads we get from 3 coin flips. 

| Outcomes | Probability |
|:--------:|:-----------:|
| HHH      | 1/8         |
| HHT      | 1/8         |
| HHT      | 1/8         |
| HTT      | 1/8         |
| THH      | 1/8         |
| THT      | 1/8         |
| TTH      | 1/8         |
| TTT      | 1/8         |

Note: [I suggest watching this video](Probability Distribution Table) if you are unclear of how these outcome and associated probabilites were calculated.

Before we go any further, let's introduce some notation that will come handy later on. 

* X = Random variable
* P(X) = Probability of X
* P(X = x) = Probability that the random variable X is equivalent to the value of x. 

# Coin Flipping Example

Let's use a coin flipping example to illustrate this in more detail. First we need to define some notation:

Let's say we perform a simple 1
In this coin flipping example, a flip of the coin can give either a head or a tail which represents our sample space. If we let our random variable X represent the number of heads 

our sample space 

we will measure the number of heads 


# Discrete Distributions: Probability Mass

* When our sample space consists of discrete outcomes (e.g. outcomes of a coin flip), then each distinct outcome has a probability associated with it. 
* When our sample space consists of continuous outcomes (e.g. diamond prices), then things get a bit more complicated. We can ask the question "what is the probability that a diamond's price is equal to \$326.52?" The answer to this question is actually quite small which might seem counter-initutive. But think about the scenario where you had 1000 diamonds and the chances of finding a diamond that was priced at exactly \$326.52 Not \$326.51 or \$326.53, but exactly \$326.52 If you think of that that way, then yeah the probability is actually really really low. But if we rephrase the question as 'what is the probability of finding diamond worth at between \$320 and \$420, then this probability starts taking on meaningful values. 

```{r}
library("ggplot2")
library("dplyr")
library("cowplot")

p1 <- diamonds %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 100)

p2 <- diamonds %>%
  ggplot(aes(x = price)) +
  geom_density()

plot_grid(p1, p2, nrow = 2, align = "v")
```

The probability of a discrete outcome, such as the probabiluty of falling into an interval on a continuous scale, is referred as as a probability mass. Where mass refers to the amount of "stuff"


# References

* [Random variables | Probability and Statistics | Khan Academy](https://youtu.be/3v9w79NhsfI)
* [Doing Bayesian Data Analysis - A Tutorial with R, JAGS, and Stan](https://sites.google.com/site/doingbayesiandataanalysis/)
* [Probability Distribution Table - Intro with tossing a coin 3 times](https://www.youtube.com/watch?v=5lpqiGixDd0)
* [What is a Probability Distribution?](http://stattrek.com/probability-distributions/probability-distribution.aspx)
