---
title: "Simple Linear Regression"
date: "January 3rd, 2016"
layout: post
output: 
  html_document:
    toc: true
tags: [R, linear regression]
---

```{r echo = FALSE}
library("knitr")
#library("svglite")
#knitr::opts_chunk$set(fig.path="{{ site.url }}/assets/gmm-em/", 
#                      dev = "svglite")
```

Simple linear regression is the one of the most basic supervised learning 
approaches. You will undoubtingly encountered and did simple linear regression
in school. And little unbeknowest to you, you would have been doing machine learning!!

* Simple linear regression 




**Table of Contents**

<ul data-toc="body" data-toc-headings="h2,h3"></ul>

## Why do simple linear regression?

First, it is good to get some initution around why you would do single linear 
regression. Simple linear regression is used when you want to model the 
relationship of some feature/variable to another variable. For instance, you 
may want to understand the relationship 


Weird issue with the Adverising.csv data where the header row starts with a 
comma. We can use a simple `sed` statement to remove that starting comma.

```{bash, eval = FALSE}
wget http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv
```

```{r setup}
library("gapminder")
library("ggplot2")
library("dplyr")
library("readr")
library("tibble")

advertising_df <- 
  read.csv("Advertising.csv") %>%
  as_tibble() %>%
  remove_rownames()

```{r sales-vs-tv}
advertising_df %>%
  ggplot(aes(x = TV, y = sales)) +
  geom_point() +
  xlab("TV advertising ") +
  ylab("Number of sales")
```

From this simple plot, you can readily see a correlation between advertising on 
TV and number of sales. In other words, it would appearing as you increase your
spending on TV advertising you also see an increase in the number of sales.

## The simple linear regression model

To represent this model, we use the following mathematical notion:

$$
Y = \beta_{0}+ \beta_{1}X + \epsilon
$$

The following describes what the different notations mean:

* $Y$: This is the variable that we want to predict. This is also known as the 
  response variable and in this particular example is the number of sales
* $\beta_{0}, \beta_{1}$: These are coefficients/parameters of the model and 
  they represent the intercept and slope of the model, respectively. 
* The $X$ is 
a random variable that represents the feature/variable we want to use to 
predict the response variaThe $\epsilon$
is an error term that we will explain in mode detail below.

## How accurate were the coefficient estimates?



## R Session

```{r}
devtools::session_info()
```

## References

