---
title:  "Safely Selecting your Columns in the Tidyverse"
date: "March 1st, 2020"
layout: post
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
tags: [R, tidyverse]
---

```{r echo = FALSE}
# Keep this when you are rendering this rmarkdown not through the r-to-jekyll.R
# script
library("knitr")
opts_chunk$set(dev = c("CairoSVG"))
```

# Data Masking

* intro to data masking.

Data mas

```{r}
library("magrittr")
dplyr::select(mtcars, mpg) %>%
  head()

dplyr::select(mtcars, mpg, carb) %>%
  head()
```

# Selecting a Single Column

My recommendation is to quote the column(s) you want to select:

```{r}
dplyr::select(mtcars, "mpg") 
```

```{r}
dplyr::select(mtcars, "mpg", "carb")
```

This ensures that you are explicitly selecting for columns in the dataframe and 
this would throw an error else wise:

```{r}
dplyr::select(mtcars, "mpg", "tiny")
```


# Safely Selecting Multiple Columns

What happens if you want to select for multiple columns with a single variable?
You can create a character vector storing the different column names:

```{r}
cols_to_select <- c("mpg", "carb")
dplyr::select(mtcars, cols_to_select)
```

Here the `dplyr::select` function is smart enough to figure out that you are
asking for multiple columns stored in the `cols_to_select` variable. However,
what happens 

Due to data masking, the "gear" column name takes precedence over the values in
your `gear` variable. This ends up resulting in an unexpected result. The 
solution in this case is to use the `all_of` function:

```{r}
gear <- c("mpg", "carb")
dplyr::select(mtcars, all_of(gear))
```

The function means to literally select "all of" of the columns that are stored 
in the character vector variable. By using `all_of`, you guard yourself against
any data masking.







# Conclusions

```{r}
devtools::session_info()
```

# References

* [Interactivity and Programming in the Tidyverse - Lionel Henry](https://resources.rstudio.com/rstudio-conf-2020/interactivity-and-programming-in-the-tidyverse-lionel-henry)
