---
title: "Safely Selecting Data Frame Columns in Your Tidyverse Code"
date: "May 8th, 2020"
layout: post
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
tags: [R, tidyverse]
---

```{r echo = FALSE}
# Keep this when you are rendering this rmarkdown not through the r-to-jekyll.R
# script
library("knitr")
opts_chunk$set(dev = c("CairoSVG"))
```

In my previous post ["Use of the .data and .env Pronouns to Disambiguate Your Tidyverse Code"]({% post_url 2020-03-01-use-data-env-pronouns-tidyverse %}), 
I discussed how using the `.data` and `.env` pronouns should be used to write 
production-grade R code. The post was inspired by Lionel Henry's talk titled 
["Interactivity and Programming in the tidyverse"](https://resources.rstudio.com/rstudio-conf-2020/interactivity-and-programming-in-the-tidyverse-lionel-henry).

In that same talk, the `all_of` function was introduced as a method to safely 
select column names from data frames. In this post, I will discuss how one should
use this function and the `.data` pronoun to safely select column names in
production-grade R code.

# Setup

First, I will do some setting up of my environment for the rest of the post:

```{r}
# Set mtcars to tibble to control the maximum number of printed rows. This is
# just to make the post easier to follow
library("magrittr")
library("tibble")
options(tibble.print_max = 6, tibble.print_min = 6)
mtcars <- as_tibble(mtcars)
```

# Safely Selecting a Single Column

As I mentioned from my last post, "data masking" enables you to "blend data with 
the workspace". This allows for you to easily select a single column from a 
data frame like this:

```{r}
dplyr::select(mtcars, mpg)
```

Here the `dplyr::select` function knows you are trying to select the "mpg" 
column from the `mtcars` data frame. If you use a variable to store the column 
name you want to select:

```{r}
col_to_select <- "mpg"
dplyr::select(mtcars, col_to_select)
```

**N.B: We'll ignore the note regarding the usage of `all_of` for now and come back
to this at the end of the post.**

The `dplyr::select` figures out you are selecting the column name that is stored
in the `col_to_select` variable and not a phantom column called "col_to_select"
(since it doesn't exist). Now what happens when you do something like this?

```{r}
gear <- "mpg"
dplyr::select(mtcars, gear)
```

Due to data masking, the "gear" column name takes precedence over the value 
stored in the `gear` variable. There are two ways to deal with this. You can 
explicitly refer to the column name by double quoting the column name.

```{r}
gear <- "mpg"

# Ignores the `gear` variable.
dplyr::select(mtcars, "gear")
```

If you want to use a variable to store the column name, then the safest thing to 
do is to combine it with the `.data` pronoun:

```{r}
gear <- "mpg"
dplyr::select(mtcars, .data[[gear]])
```

# Safely Selecting Multiple Columns

If you want to safely select multiple columns, then you should start by double
quoting the columns names to make it explicit:

```{r}
dplyr::select(mtcars, "mpg", "carb")
```

If you want to use a variable to select multiple columns, you should create a 
character vector storing the different column names: 

```{r message = FALSE}
cols_to_select <- c("mpg", "carb")
dplyr::select(mtcars, cols_to_select)
```

Here the `dplyr::select` function is smart enough to figure out that you are
asking for multiple columns stored in the `cols_to_select` variable. However,
what happens if you have something like this?

```{r}
gear <- c("mpg", "carb")
dplyr::select(mtcars, gear)
```

Similar to the scenario where we were selecting a single column, data masking
results in the "gear" column name taking precedence over the values in the 
`gear` variable. This ends up creating an unexpected result. This is where the
`all_of` function comes in:

```{r}
gear <- c("mpg", "carb")
dplyr::select(mtcars, all_of(gear))
```

The function means to literally select "all of" the columns that are stored in 
the character vector variable. By using `all_of`, you guard yourself against
any data masking.

# `.data` vs. `all_of`

Near the beginning of the post, we saw the following note when using a variable
with the `dplyr::select` function:

```{r}
col_to_select <- "mpg"
dplyr::select(mtcars, col_to_select)
```

Based on this, it seems that using the function `all_of` is the recommended way 
to select columns when using a variable even when it's only a single column. 
In other words,

```{r}
col_to_select <- "mpg"
dplyr::select(mtcars, all_of(col_to_select))
```

This is as valid as using the `.data` pronoun. My personal preference is to 
stick to using the `.data` pronoun when I only want to select a single column
and use `all_of` when my character vector has multiple values in it. This simply
makes it more obvious to me what I should be expecting from `dplyr::select` 
call.

# Conclusions

In conclusion, my recommendations for safely selecting data frame columns in the 
tidyverse are as follows:

* Be explicit in the single and multiple column selections by double quoting the 
    column names.
* If you want to use a variable to store a single column name, then combine it 
    with the `.data` pronoun.
* If you want to use a variable to store a character vector of multiple column 
    names, then use the `all_of` function.

```{r}
devtools::session_info()
```

# References

* [Interactivity and Programming in the Tidyverse - Lionel Henry](https://resources.rstudio.com/rstudio-conf-2020/interactivity-and-programming-in-the-tidyverse-lionel-henry)
